---
title: "Embedding Regression with conText"
output:
  rmarkdown::github_document:
    math_method: webtex
vignette: >
  %\VignetteIndexEntry{Embedding Regression with conText}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = FALSE,
                      dpi = 72,
                      warning = FALSE,
                      message = FALSE,
                      comment = "##",
                      fig.width = 6,
                      fig.height = 4,
                      tidy = FALSE)
```

This tutorial walks through the **embedding regression** workflow in **conText** end
to end: from a tokenized corpus to a fitted model, to reading and visualizing the
results, predicting embeddings for covariate profiles, and tracing how a term's
usage changes along a continuous covariate. It complements the
[Quick Start Guide](quickstart.md), which focuses on the descriptive ("nearest
neighbours") tools.

We use the small objects bundled with the package -- `cr_sample_corpus`,
`cr_glove_subset` and `cr_transform` -- so everything here is reproducible. They are
tiny, so treat the numbers as illustrative only.

```{r}
library(conText)
library(quanteda)
library(dplyr)
library(ggplot2)
```

# What embedding regression does

For a focal term, conText represents **each instance** of that term by an
"*a la carte*" (ALC) embedding: the average of the pre-trained embeddings of the
words in its context, multiplied by a transformation matrix `A`. Stacking those
per-instance embeddings as the outcome `Y`, conText fits

$$\mathbf{Y} = \mathbf{X}\boldsymbol{\beta} + \mathbf{E},$$

a multivariate regression of the embeddings on document-level covariates `X`. Each
coefficient `β` is itself a `D`-dimensional vector; its (debiased, squared)
**norm** measures how much that covariate shifts the term's usage, and conText
attaches inference to those norms (a permutation test for p-values and, optionally,
a jackknife for confidence intervals).

# Fitting a model

`conText()` takes a formula with the focal term on the left-hand side and the
covariates on the right, plus a tokenized corpus and the pre-trained embeddings /
transformation matrix.

```{r}
toks <- tokens(cr_sample_corpus)

set.seed(2021L)
model <- conText(immigration ~ party + gender,
                 data = toks,
                 pre_trained = cr_glove_subset,
                 transform = TRUE, transform_matrix = cr_transform,
                 jackknife = FALSE,        # set TRUE to add confidence intervals (slower)
                 permute = TRUE, num_permutations = 100,
                 verbose = FALSE)
```

Character/factor covariates are automatically turned into `0/1` indicator
variables, leaving out a base category. Here `party` becomes `party_R` (base =
Democrat) and `gender` becomes `gender_M` (base = Female).

# Reading the results

The fitted object is a matrix of coefficients (one row per covariate, including the
intercept). Three methods make it easy to inspect:

`summary()` prints the model dimensions and the normed-coefficient table:

```{r}
summary(model)
```

`tidy()` returns that table as a tibble (handy for further manipulation or
plotting):

```{r}
tidy(model)
```

`plot()` shows the deflated norm of each coefficient. Coefficients significant at
the 0.05 level (using the permutation p-values) are marked with a `*`; when the
model is fit with `jackknife = TRUE`, confidence intervals are drawn as well.

```{r}
plot(model)
```

# Interaction effects

Interactions use the usual formula syntax. `party*gender` expands to
`party + gender + party:gender`; the interaction enters as the product of the
corresponding indicator columns and gets its own coefficient and test.

```{r}
set.seed(2021L)
model_int <- conText(immigration ~ party * gender,
                     data = toks,
                     pre_trained = cr_glove_subset,
                     transform = TRUE, transform_matrix = cr_transform,
                     jackknife = FALSE, permute = TRUE, num_permutations = 100,
                     verbose = FALSE)

rownames(model_int)
```

# Predicting embeddings for covariate profiles

The coefficients combine into the ALC embedding for any covariate profile. Rather
than adding rows by hand, use `predict()` with a `newdata` of profiles (columns are
the coefficient names; interaction columns are filled automatically from their
components).

```{r}
group_wvs <- predict(model_int,
                     newdata = data.frame(
                       party_R  = c(0, 0, 1, 1),
                       gender_M = c(0, 1, 0, 1),
                       row.names = c("Dem-Female", "Dem-Male", "Rep-Female", "Rep-Male")))
dim(group_wvs)
```

The result is an ordinary matrix (one row per profile) that feeds directly into the
descriptive tools. For example, the nearest neighbours of each group's embedding:

```{r}
nns(group_wvs, N = 5, pre_trained = cr_glove_subset, as_list = FALSE)
```

# Continuous covariates: usage along a dimension

`predict()` also accepts continuous covariates, so you can trace how a term's usage
moves along a scale -- for instance the first dimension of DW-NOMINATE
(`nominate_dim1`), included in the sample corpus.

```{r}
set.seed(2021L)
model_nom <- conText(immigration ~ nominate_dim1,
                     data = toks,
                     pre_trained = cr_glove_subset,
                     transform = TRUE, transform_matrix = cr_transform,
                     jackknife = FALSE, permute = TRUE, num_permutations = 100,
                     verbose = FALSE)
```

Predict the ALC embedding at a grid of NOMINATE values, then measure its cosine
similarity to a few features of interest:

```{r}
grid <- seq(-1, 1, by = 0.2)
nom_wvs <- predict(model_nom, newdata = data.frame(nominate_dim1 = grid,
                                                   row.names = as.character(grid)))

effects <- cos_sim(nom_wvs, pre_trained = cr_glove_subset,
                   features = c("reform", "enforce"), as_list = FALSE)
head(effects)
```

Because `cos_sim()` returns a tidy data frame, plotting the "effect" of the
covariate is a one-liner:

```{r}
effects %>%
  mutate(nominate = as.numeric(target)) %>%
  ggplot(aes(nominate, value, color = feature)) +
  geom_line() + geom_point() +
  labs(x = "DW-NOMINATE (dim 1)", y = "cosine similarity to feature",
       color = NULL) +
  theme_bw()
```

# Exploratory tools with uncertainty

Alongside the regression, the `get_*()` wrappers go from a tokenized corpus to
group-level nearest neighbours / similarities, with bootstrap standard errors and
confidence intervals. The intervals are standard percentile-bootstrap intervals at
the requested `confidence_level` (e.g. a 95% interval uses the 2.5th and 97.5th
percentiles).

```{r}
immig_toks <- tokens_context(toks, pattern = "immigration", window = 6L, verbose = FALSE)

set.seed(2021L)
immig_nns <- get_nns(x = immig_toks, N = 5,
                     groups = docvars(immig_toks, "party"),
                     candidates = character(0),
                     pre_trained = cr_glove_subset,
                     transform = TRUE, transform_matrix = cr_transform,
                     bootstrap = TRUE, num_bootstraps = 100,
                     confidence_level = 0.95,
                     as_list = FALSE)
head(immig_nns)
```

# Notes on inference and performance

* **Test statistic.** conText tests the *debiased* squared norm of each coefficient
  (`normed.estimate.deflated`), which subtracts an estimate of the sampling-noise
  contribution from the raw norm (Green et al. 2025). The raw norm
  (`normed.estimate.orig`) is also reported.
* **p-values vs. intervals.** `permute = TRUE` gives empirical p-values via a
  permutation test; `jackknife = TRUE` adds standard errors and confidence
  intervals.
* **Clustering.** Pass `cluster_variable` to cluster the standard errors (e.g. by
  speaker), so repeated instances from the same unit are not treated as independent.
* **Performance.** The jackknife is leave-one-out, so on large corpora it can be
  slow; use `jackknife_fraction` to subsample, or `parallel = TRUE` with a
  registered backend.

# References

Rodriguez, P. L., Spirling, A., and Stewart, B. M. (2023). Embedding Regression:
Models for Context-Specific Description and Inference. *American Political Science
Review*, 117(4), 1255-1274.

Khodak, M., Saunshi, N., Liang, Y., Ma, T., Stewart, B., and Arora, S. (2018). A La
Carte Embedding: Cheap but Effective Induction of Semantic Feature Vectors. *ACL*.
